A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-07-02 paper/paper.tex
paper: more progress

\subsection*{Fibered vs indexed presentation of semi-simplicial and semi-cubical sets}
A family of sets can commonly be represented in two ways: as a family properly speaking, indexed over a set or as a set together with a map specifying for each element of the set on which index it depends. In the latter case, the set associated to a given index is the fiber, i.e. inverse image, of the index under the map. The two presentations are equivalent, what can be formally stated in a type-theoretic language as a fibered/indexed correspondence~\footnote{At a discussion on the Zulip channel for the HoTT community, alternative names fibration/family correspondence and unbundled/bundled correspondence were proposed, among others. In our experience, the opposition fibered/indexed is spontaneously understood by many listeners. In particular, it echoes to the Grothendieck construction of fibered categories from indexed categories. The most elementary instance of the correspondence, with $\Type$ instead of $\U$, that is $(\Sigma S: \Type. (S \rightarrow T)) \simeq (T \rightarrow \Type)$, is sometimes called ``Grothendieck construction for dummies'' (its proof requires univalence, see~\cite{hottbook}).}:
\begin{equation*}
  \mbox{(fibered)}\qquad(\Sigma S: \U. (S \rightarrow T)) \simeq (T \rightarrow \U)\qquad \mbox{(indexed)}
\end{equation*}
where $\U$ represents the subset of types with a unique proof of equality, that is the type-theoretic counterpart of the category of sets (within some universe).

While a presheaf over an index category, i.e. a family of sets connected with maps according to the morphisms of the index category, can be seen as living on the indexed side of the correspondence, opposing to discrete Grothendieck fibrations on the fibered side, there are situations where a presheaf $F$ can also be seen as living on the fibered side of the correspondence. Indeed, when the index category is direct, i.e. with a downwards-well-founded collection of non-identity morphisms, as is the case of semi-simplicial and semi-cubical sets, each $F(X)$ serving as source of a map $F(X) \stackrel{F(f)}{\rightarrow} F(Y)$ can be turned into a family $F(X)$ indexed on $F(Y)$, and, since the maps are well-founded downward, this can be iterated from below, eventually producing a collection of families which we call an \emph{indexed presentation} of the presheaf over the direct category.

For instance, in the case of a semi-cubical set~\citep{grandis03,buchholtz17}, presented with $2n$ face maps from the set of $n$-cubes to the set of $(n-1)$-cubes, the standard presheaf approach prescribes a family of sets fibered the ones over the others (up to coherence conditions such). Formulated in type theory, it corresponds to:

@@

On the other side, the indexed presentation of the same presheaf is a stratified sequence of families indexed on elements of families of lower rank, taking into account the coherence conditions to prevent duplications. Formulated in type theory, it corresponds to:

@@

The indexed definition of a presheaf over a direct category is technically more involved than the presheaf definition as it requires hard-wiring in the structure the dependencies between elements of the sets of the presheaf, including the coherence conditions between these dependencies, such as taking the $i$-th face of the $j$-th face of a $n$-simplex is the same as taking the $(j-1)$-th face of the $i$-th face (when $j>i$). On the other side, defining a concrete presheaf in indexed form only requires to provide the family since the responsibility of defining maps and showing the coherence conditions is already accounted for in the definition, what is not the case by default in the presheaf presentation. In category-theoretic terms, the indexation set of the family $X_n$ in the indexed presentation is called a \emph{matching object}. The indexed presentation of presheaves over a direct category can then be identified with a subclass of ordinary presheaves known as Reedy fibrant where Reedy fibrancy precisely expresses that the set $X_n$ in the fibered presentation is a $\Sigma$-type over an instance up to isomorphism of the matching object at level $n$~\citep{shulman14,kraus17}.

@@

\subsection*{\texorpdfstring{$\nu$}{ν}-sets}
Let us call $\nu$-sets, the generalization of augmented semi-simplicial sets and semi-cubical sets. To obtain this, we extend $\DeltaPlus$ and $\Cube$ in a straightforward manner into a category which we call $\nu$-semi-shape category. The morphisms of the $\nu$-semi-shape category are sequences of elements of a set $\nu$ of arbitrary cardinal, extended with $\kstar$, so that the following holds.

@@

  \begin{figure}[H]
    \centering
    \begin{align*}
      \Obj(\Cube) & := \Nat                                                                                \\
      \Hom(p, n)  & := \{l \in [L, R, \kstar]^n \mid \text{number of $\kstar$ in $l = p$}\}                \\
      g \circ f   & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                              \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = L$ or $R$}                          \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = L$, $R$, or $\kstar$} \\
      \end{cases} \\
      \id         & := [\kstar, \ldots, \kstar] \text{ $n$ times}
    \end{align*}
  \end{figure}

@@

>>>>

\subsection*{Fibered vs indexed presentation of semi-simplicial and semi-cubical sets}
A family of sets can commonly be represented in two ways: as a family properly speaking, indexed over a set $S$, or as a set $T$ together with a map specifying for each element of the set $T$ on which index it depends in $S$. In the first case, we call it an \emph{indexed} presentation. In the latter case, the set associated to a given index is the fiber, i.e. inverse image, of the index under the map, so we call it a \emph{fibered} presentation. The two presentations are equivalent and the equivalence can be phrased concisely in the language of homotopy type theory~\cite{hottbook} as a fibered/indexed correspondence\footnote{At a discussion on the Zulip channel for the HoTT community, alternative names fibration/family correspondence and unbundled/bundled correspondence were proposed, among others. In our experience, the opposition fibered/indexed is spontaneously understood by many listeners. In particular, it echoes to the Grothendieck construction of fibered categories from indexed categories. The most elementary instance of the correspondence, with $\Type$ instead of $\U$, that is $(\Sigma S: \Type. (S \rightarrow T)) \simeq (T \rightarrow \Type)$, is sometimes called ``Grothendieck construction for dummies'' (its proof requires univalence, see~\cite{hottbook}).}
\begin{equation*}
  \mbox{(fibered)}\qquad(\Sigma S: \U. (S \rightarrow T)) ~\simeq~ (T \rightarrow \U)\qquad \mbox{(indexed)}
\end{equation*}
where $\U$ represents sets in homotopy type theory as the subset of types with a unique proof of equality (within some universe).

A \emph{presheaf} on an category is a family of sets connected with maps according to the morphisms of the category. As such, it lives on the indexed side of the correspondence, opposing on the fibered side to what is called \emph{discrete Grothendieck fibrations}. However, there are situations where a presheaf $F$ can also be seen as living on the fibered side of the correspondence. Indeed, when the indexing category is direct, i.e. with a downwards-well-founded collection of non-identity morphisms, as is the case of semi-simplicial and semi-cubical sets, each $F(X)$ serving as source of a map $F(X) \stackrel{F(f)}{\rightarrow} F(Y)$ can be turned into a family $F(X)$ indexed on $F(Y)$, and, since the maps are well-founded downward, this can be iterated from below, eventually producing a collection of families which we call an \emph{indexed presentation} of the presheaf over the direct category.

For instance, in the case of a semi-cubical set~\citep{grandis03,buchholtz17}, presented with $2n$ face maps from the set of $n$-cubes to the set of $(n-1)$-cubes, the standard presheaf approach prescribes a family of sets fibered the ones over the others (up to cubical faces identities). Formulated in type theory, it corresponds to:

@@

On the other side, the indexed presentation of the same presheaf is a stratified sequence of families indexed on elements of families of lower rank, taking into account the coherence conditions to prevent duplications. Formulated in type theory, it takes the form:

@@

The indexed definition of a presheaf over a direct category is technically more involved than the presheaf definition as it requires hard-wiring in the structure the dependencies between elements of the sets of the presheaf, including the coherence conditions between these dependencies, such as taking the $i$-th face of the $j$-th face of a $n$-simplex is the same as taking the $(j-1)$-th face of the $i$-th face (when $j>i$). On the other side, defining a concrete presheaf in indexed form only requires to provide the family since the responsibility of defining maps and showing the coherence conditions is already accounted for in the definition, what is not the case by default in the (ordinary) presheaf presentation.

In category-theoretic terms, the indexi set of the family $X_n$ in the indexed presentation is called a \emph{matching object}. The indexed presentation of presheaves over a direct category can then be identified with a subclass of ordinary presheaves known as Reedy fibrant where Reedy fibrancy precisely expresses that the set $X_n$ in the fibered presentation is a $\Sigma$-type over an instance up to isomorphism of the matching object at level $n$~\citep{shulman14,kraus17}.

@@

@@

  \begin{figure}[H]
    \centering
    \begin{align*}
      \Obj(\Cube) & := \Nat                                                                                 \\
      \Hom(p, n)  & := \{l \in [L, R, \kstar]^n \mid \text{number of $\kstar$ in $l = p$}\}                 \\
      g \circ f   & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                               \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = L$ or $R$}                           \\
        a :: (g' \circ f') & \text{if $g = (\kstar :: g')$, $f = (a :: f')$, where $a = L$, $R$, or $\kstar$} \\
      \end{cases} \\
      \id         & := [\kstar, \ldots, \kstar] \text{ $n$ times}
    \end{align*}
  \end{figure}

@@

\subsection*{\texorpdfstring{$\nu$}{ν}-sets}
Let us call $\nu$-sets, the generalization of augmented semi-simplicial sets and semi-cubical sets. To obtain this, we extend $\DeltaPlus$ and $\Cube$ in a straightforward manner into a category which we call $\nu$-semi-shape category.

\begin{definition}[$\nu$-semi-shape categroy]
  The definition of $\nu$-semi-shape category is shown below. Note that, if $g \circ f$ is well-defined, then the length of $f$ is less than that of $g$. It can be shown that composition is associative and that $\id$ is neutral.

  \begin{figure}[H]
    \centering
    \begin{align*}
      \Obj       & := \Nat                                                                            \\
      \Hom(p, n) & := \{l \in (\nu \cup \kstar) \mid \text{number of $\kstar$ in $l = p$}\}           \\
      g \circ f  & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                         \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $R \in \nu$}                        \\
        a :: (g' \circ f') & \text{if $g = (\kstar :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
      \end{cases} \\
      \id        & := [\kstar, \ldots, \kstar] \text{ $n$ times for $\id \in \Hom(n, n)$}
    \end{align*}
  \end{figure}
\end{definition}

The morphisms of the $\nu$-semi-shape category are sequences of elements of a set $\nu$ of arbitrary cardinal, extended with $\kstar$:

so that the following holds.
